Different forms of uncertainty differently affect the evolution of social learning

### 2022-08-13 ms.tex
added stubs for sensitivity and did minor revisions

\emph{Agents} here are autonomous, intelligent, simulated problem solvers. Each
agent evaluates its environment and chooses how to act within that environment based
on a set of empirically-motivated rules.  Our agents select which behavior to do
probabilistically, weighted by the softmax function applied to the mean payoff they
have observed for each behavior. The softmax function is a biologically plausible
generalization of behavior selection under uncertainty that enables agents to often
greedily exploit the most lucrative behavior (based on their observations) but also
to sometimes explore alternatives~\cite{Schulz2019,Collins2013,Daw2006,Yechiam2005}.
Such individual-level intelligence is a critical adaptation that has apparently
co-evolved with the capacity for social learning, and has a marked influence on the
cultural evolution of social learning as we will demonstrate in our Analysis.
We will see that individual-level intelligence allows agents to recover from
the receipt of outdated information due to temporal environmental variability.

In \emph{evolutionary} agent-based models like ours, agent life cycles are
programmed to cause agents to reproduce and die off, where reproducers pass on
heritable traits to their offspring. 
Natural selection occurs in the model when agents with higher
payoffs preferentially reproduce. This is a model of cultural evolution, so
``reproduction'' means teaching the meta-cognitive strategy of whether one should learn
socially at all; ``death'', in this interpretation, means
loss of the potential to socially influence others. 
This sets up an inter-generational structure
that we also use as a scaffold for social learning: we assume social learning occurs
\emph{obliquely}, meaning learning happens when a new
generation of agents each select a teacher from the previous generation, possibly
including their parents.  
Evolution, then, acts as an optimization routine to identify which heritable trait
values are most rewarded under in different environmental uncertainty contexts,
defined by different uncertainty parameter settings
(Table~\ref{tab:uncertaintyParameters}).

Agents have one heritable trait in this model, whether or not
an agent, $i$, learns socially, denoted $s_i$. $s_i$ is binary: it is 1
if the agent is a social learner, and 0 otherwise. In our analysis, we 
aggregate $s_i$ over all agents and simulation trials to understand the affects of
different uncertainty parameter settings. 
Our computational analysis amounts to observing patterns in
the aggregated $s_i$, supported by measuring the number of generations 
to model convergence and expected payoffs. 
To understand how different varieties of uncertainty interact
to affect the evolution of social learning, we systematically created an array of
simulated environments by systematically varying the uncertainty parameters.

@@

We performed sensitivity analyses to ensure that our main analysis is reasonably
robust to a range of auxiliary parameters. Specifically,
we varied the population size, $N$, to explore whether smaller $N$ would 
show more drift due to finite population size effects, but also generally support
our main conclusions. Next, we varied the
softmax greediness parameter $\beta \in \{1, 100\}$ to supplement the main results
that used $\beta = 10$. We do not
expect agents with different $\beta$ to perform equally well 
individually, so certain $\beta$ values may themselves suppress social learning
by undermining individual learning. Nonetheless, our results
should be valid over some range of $\beta$. Finally, we varied the
number of prospective teachers $N_T \in \{2, 10, 20\}$ 
to supplement the main results that used $N_T = 5$. The number of prospective
teachers should not change whether social learning is optimal, 
but it may induce more drift since the benefit of social learning may
be more difficult to detect with fewer prosopective teachers.

\subsubsection{Implementation}
Our model was implemented in the Julia programming language~\cite{Bezanson2017} 
using the Agents.jl agent-based modeling library~\cite{Datseris2022} and run
on the Sherlock supercomputing cluster at Stanford University. Model code and
data are publicly available on GitHub\footnote{\url{https://github.com/mt-digital/UncMod}}.

@@

Increased selection set size $B$ led nearly uniformly to more prevalent social
learning, with all other uncertainty parameters held constant. This is first 
because it is more difficuilt to find the optimal behavior with a larger $B$.
% more options present a more difficult problem for trial-and-error learning within
% a lifespan. 
This effect is amplified because agents in a world with larger $B$ are more
cognitively flexible compared to those in a smaller $B$ world. 
When $B$ is larger, agents are relatively more likely to choose an alternative to
their current known optimum. %is easier to correct for larger $B$ relative to smaller $B$.  
To understand this, consider the case where a social learner child learns
that behavior 1 has an expected payoff of 0.8, and the rest have an expected
payoff of 0, but the optimal behavior changed between generations. Assume
the child tries behavior 1, but it pays off 0.
If $B=2$, then at the next within-generation time step, $\ledger = (0.4, 0.0)$. The probability
the agent explores behavior 2 is now 0.02. However, if $B=10$ the agent has
$\ledger = (0.4, 0.0, \ldots, 0.0)$; the probability
the agent explores one of the nine alternative behaviors increases to 0.14.
If the agent again tries behavior 1 and it pays 0, the probability of exploring alternatives
with $B=2$ is 0.21, but with $B=10$ the probability of exploring alternatives
greatly increases to 0.71. \mt{I double checked these calculations, but would appreciate if
someone triple checked it}. This explanation does not apply when $L=1$, where
individual learning does not occur.

@@

\mt{These here used $N=100$; need to remake with $N=1000$}

\begin{table}[h] \caption{Max. iterations = 20000.} \label{tab:convergence} \centering
  \begin{tabular}{cccc} \toprule $B$ & \# not at fixation & \# total series & Pct. not fixated
    \\ \midrule  2  & 0  & 132000 & 0.0 \% \\ 4  & 0  & 132000 & 0.0 \% \\ 10 & 44 & 132000 &
    0.00033  \% \\ \bottomrule \end{tabular} \end{table}

@@

\begin{figure}
  % \addtocounter{figure}{-1}
  \caption{Comparison of expected payoffs from our computational anaysis (solid
    lines with circles), expected individual payoffs with no social learning
    (dashes without markers), and expected payoffs for all-social learner 
    populations. When $\pilow = 0.8$, all-social learner populations underperform
    since social learners are not subject to an initial evolutionary bottleneck
    due to the presence of asocial learners, and behavioral misinformation
    is allowed to circulate more freely than in the base case tested by 
    our model with an initially even distribution of social and asocial learners.
}
  \label{fig:fullMeanPrevNetPayoffs}
  \centering
    \includegraphics[width=\textwidth]{Figures/supplement/fullMeanPrevNetPayoffs.pdf}
\end{figure}

@@

\mt{These here used $N=100$; need to remake with $N=1000$}

\vspace{-3em} \begin{figure} %\addtocounter{figure}{-1} 
  \centering
  \caption{Sensitivity analysis of the main results for the softmax parameter $\beta = 100$ and
  $\beta=1$. Recall the main results were obtained with $\beta = 10$.}
  \label{fig:softmaxSensitivity} \vspace{2em}
  \begin{subfigure}{\textwidth}
	\caption{$\beta = 100$}
	\includegraphics[width=\textwidth]{Figures/supplement/sensitivity_tau=0.01/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
\newpage
\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{$\beta = 1$}
	\includegraphics[width=\textwidth]{Figures/supplement/sensitivity_tau=1.0/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

\newpage
\subsection{Population size sensitivty analysis}

\begin{figure}
  \centering
  \caption{
	Sensitivity analysis of the main results for different population
	sizes, $N=50,200,1000$. Recall $N=100$ was used to generate main 
	text results.
  }
  \label{fig:populationSensitivity}
  \begin{subfigure}{\textwidth}
	\caption{$N=50$}
	\includegraphics[width=\textwidth]{Figures/supplement/nagents=50/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

@@

\mt{These here used $N=100$; need to remake with $N=1000$}

\vspace{-3em}
\begin{figure}
  \centering
  \caption{Number of prospective teachers sensitivity analysis for $N_T=2,10,20$. Recall
  $N_T=5$ was used to generate main text results.}
  \label{fig:softmaxSensitivity}
  \vspace{2em}
  \begin{subfigure}{\textwidth}
	\caption{$N_T = 2$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=2/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
\newpage
\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{$N_T = 10$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=10/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
\newpage
\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{$N_T = 20$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=20/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

>>>>

\emph{Agents} here are autonomous, intelligent, simulated problem solvers. Each
agent evaluates its environment and chooses how to act within that environment based
on a set of empirically-motivated rules.  Our agents select which behavior to do
probabilistically, weighted by the softmax function applied to the mean payoff they
have observed for each behavior. The softmax function is a biologically plausible
generalization of behavior selection under uncertainty that enables agents to often
greedily exploit the most lucrative behavior (based on their observations) but also
to sometimes explore alternatives~\cite{Schulz2019,Collins2013,Daw2006,Yechiam2005}.
Such individual-level intelligence is a critical adaptation that has 
co-evolved with the capacity for social learning, and we will show this has a marked influence 
on the cultural evolution of social learning.

In \emph{evolutionary} agent-based models like ours, agent life cycles are
programmed to cause agents to reproduce and die off, where reproducers pass on
heritable traits to their offspring. 
Natural selection occurs in the model when agents with higher
payoffs preferentially reproduce. This is a model of cultural evolution, so
``reproduction'' means teaching the meta-cognitive strategy of whether one should learn
socially at all; ``death'', in this interpretation, means
loss of the potential to socially influence others. 
This sets up an inter-generational structure
that we also use as a scaffold for \emph{oblique} social learning, meaning children
with the social learning trait learn about payoffs from a member of the
previous generation. Evolution acts as an optimization routine to identify
whether individuals with the social learning trait will outperform those without.

Agents inherit one trait in this model, whether or not
an agent, $i$, learns socially. If $i$ has the trait we write $s_i=1$
 and 0 otherwise. In our analysis, we 
average $s_i$ over all agents and simulation trials to understand the affects of
different uncertainty parameter settings. 
Our computational analysis amounts to observing patterns in
the aggregated $s_i$, supported by measuring the number of generations 
to model convergence and expected payoffs. 
To understand how different varieties of uncertainty interact
to affect the evolution of social learning, we systematically created an array of
simulated environments by systematically varying the uncertainty parameters.

@@

We performed sensitivity analyses to ensure that our main analysis is reasonably
robust to a range of auxiliary parameters. First,
we varied the population size, $N$, to explore whether smaller $N$ would 
show more drift due to finite population size effects, but also generally support
our main conclusions. Next, we varied the
softmax greediness parameter $\beta \in \{1, 100\}$ to supplement the main results
that used $\beta = 10$. We do not
expect agents with different $\beta$ to perform equally well 
individually, so certain $\beta$ values may themselves suppress social learning
by undermining individual learning. Nonetheless, our results
should be valid over some range of $\beta$. Finally, we varied the
number of prospective teachers $N_T \in \{2, 10, 20\}$ 
to supplement the main results that used $N_T = 5$. The number of prospective
teachers should not change whether social learning is optimal, 
but it may induce more drift since the benefit of social learning may
be more difficult to detect with fewer prosopective teachers.

\subsubsection{Implementation}

Our model was implemented in the Julia programming language~\cite{Bezanson2017} 
using the Agents.jl agent-based modeling library~\cite{Datseris2022} and run
on the Sherlock supercomputing cluster at Stanford University. Model code and
data are publicly available on GitHub\footnote{\url{https://github.com/mt-digital/UncMod}}.

@@

Increased selection set size $B$ led nearly uniformly to more prevalent social
learning, with all other uncertainty parameters held constant. This is because,
first, it is more difficuilt to find the optimal behavior with a larger $B$.
% more options present a more difficult problem for trial-and-error learning within
% a lifespan. 
Second, this effect is amplified because agents in a world with larger $B$ are more
cognitively flexible compared to those in a smaller $B$ world. 
When $B$ is larger, agents are relatively more likely to choose an alternative to
their current known optimum. %is easier to correct for larger $B$ relative to smaller $B$.  
To understand this, consider the case where a social learner child learns
that behavior 1 has an expected payoff of 0.8, and the rest have an expected
payoff of 0, but the optimal behavior changed between generations. Assume
the child tries behavior 1, but it pays off 0.
If $B=2$, then at the next within-generation time step, $\ledger = (0.4, 0.0)$. The probability
the agent explores behavior 2 is now 0.02. However, if $B=10$ the agent has
$\ledger = (0.4, 0.0, \ldots, 0.0)$; the probability
the agent explores one of the nine alternative behaviors increases to 0.14.
If the agent again tries behavior 1 and it pays 0, the probability of exploring alternatives
with $B=2$ is 0.21, but with $B=10$ the probability of exploring alternatives
greatly increases to 0.71. \mt{I double checked these calculations, but would appreciate if
someone triple checked it}. This explanation does not apply when $L=1$, where
individual learning does not occur.

@@

\begin{table}[h] \caption{Max. iterations = 20000.} \label{tab:convergence} \centering
  \begin{tabular}{cccc} 
    \toprule $B$ & \# not at fixation (out of 132k) & \% not fixated \\ 
    \midrule  2  & 192  & 0.1 \% \\ 
              4  & 274  & 0.2 \% \\ 
              10 & 2935 & 2.0 \% \\ 
    \bottomrule \end{tabular} 
\end{table}

@@

\begin{figure}
  \caption{Comparison of expected payoffs from our computational anaysis (solid
    lines with circles), expected individual payoffs with no social learning
    (dashes without markers), and expected payoffs for all-social learner 
    populations. When $\pilow = 0.8$, all-social learner populations underperform
    since social learners are not subject to an initial evolutionary bottleneck
    due to the presence of asocial learners, and behavioral misinformation
    is allowed to circulate more freely than in the base case tested by 
    our model with an initially even distribution of social and asocial learners.
}
  \label{fig:fullMeanPrevNetPayoffs}
  \centering
    \includegraphics[width=\textwidth]{Figures/supplement/fullMeanPrevNetPayoffs.pdf}
\end{figure}

@@

We tested two additional settings of the softmax behavior selection greediness parameter,
$\beta=1,100$, which reproduced qualitative patterns in the evolution of social learning. In
the main text we used $\beta=10$.  Softmax greediness is achieved formally by increasing the
probability mass for selecting behaviors that have paid off more frequently over that agent's
life history. In other words, greater $\beta$ increases the probabilty that the optimal
observed behavior is chosen over all the rest. If $\beta$ is too small, agents will fail to
fully exploit their knowledge about which behavior is optimal because they too frequently
explore less profitable alternatives.  If $\beta$ is too great the agents could more regularly
settle on a sub-optimal behavior because it paid off first.  In a fuller evolutionary
exploration of model parameters it would make sense for $\beta$ to co-evolve with the social
learner trait.  However, our sensitivity analysis demonstrates that our main conclusions of how
social learning evolves under uncertainty remain unchanged for different $\beta$ over three
orders of magnitude.

When $\beta=1$, most general trends are preserved, but in many cases the evolution
of social learning is suppressed or drift dominates due to 
due to the overall poor quality of information acquired through softmax search-based
individual learning.

When $\beta=100$, there

Other smaller details can also be observed between tested $\beta$. 
First, there is generally more drift when $\beta=1$, 
which can be seen in certain $(\pilow, B)$ pairs
(entries in plot matrices in Figure~\ref{fig:softmaxSensitivity}a). 
For example, when $\pilow = 0.1$ and $B=10$, there are a few more 
data points with greater drift, i.e., where $\meansl \approx 0.5$. 
The variation in our results when we set $\beta=100$
(Figure~\ref{fig:softmaxSensitivity}b), changed from the
main text where $\beta=10$, are more varied and subtle compared
to setting $\beta=1$. In some cases, such as $(\pilow=0.1, B=10)$ 
and $(\pilow=0.45, B=2)$,
there appears to be more drift across all $L$ around the transition
point when $\meansl = 1$ goes to 0 as $u$ increases. However it appears
that there is more drift at the transition for $\beta=10$ for
$(\pilow=0.45, B=10)$.

\vspace{-3em} \begin{figure} %\addtocounter{figure}{-1} 
  \centering
  \caption{Sensitivity analysis of the main results for the softmax parameter $\beta = 100$ and
  $\beta=1$. Recall the main results were obtained with $\beta = 10$.}
  \label{fig:softmaxSensitivity} \vspace{2em}
  \begin{subfigure}{\textwidth}
	\caption{$\beta = 1$}
	\includegraphics[width=\textwidth]{Figures/supplement/sensitivity_tau=1.0/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
\newpage
\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{$\beta = 100$}
	\includegraphics[width=\textwidth]{Figures/supplement/sensitivity_tau=0.01/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

\subsection{Population size sensitivty analysis}

Drift is greater for smaller populations sizes, especially when
$\pilow=0.8$, but patterns generally remain.

\begin{figure}
  \centering
  \addtocounter{figure}{-1}
  \caption{
	Sensitivity analysis of the main results for different population
	sizes, $N=50,200,1000$. Recall $N=100$ was used to generate main 
	text results.
  }
  \label{fig:populationSensitivity}
  \begin{subfigure}{\textwidth}
	\caption{$N=50$}
	\includegraphics[width=\textwidth]{Figures/supplement/nagents=50/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

@@

Slightly more drift when $N_T = 2$, slightly less drift when $N_T = 20$, 
but mostly just reproduces results for main text setting $N_T = 5$.

\vspace{-3em}
\begin{figure}
  \centering
  \addtocounter{figure}{-1}
  \caption{Number of prospective teachers sensitivity analysis for $N_T=2,10,20$. Recall
  $N_T=5$ was used to generate main text results.}
  \label{fig:softmaxSensitivity}
  \vspace{2em}
  \begin{subfigure}{\textwidth}
	\caption{$N_T = 2$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=2/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
\newpage
\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{$N_T = 10$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=10/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
\newpage
\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{$N_T = 20$}
	\includegraphics[width=\textwidth]{Figures/supplement/nteachers=20/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
